---
title: "R Fundamentals"
Student: Channing Powers
output:
  html_document:
    df_print: paged
Instructor: Bogdan Anastasiei
---

#################################
# CREATING VECTORS WITH THE c() FUNCTION
#################################

### a vector is a sequence of elements of the same type:
### numeric, character (string) or logical
### NUMERICAL | CHARACTER | LOGICAL


### one of the most common ways to create vectors
### is using the c() function
### c stands for concatenate
### c() concatenates the values to create a vector.
### The values are the arguments of the function.
### Values in a vector can be Numeric,Categorical, or both.

### Numeric
```{r}
x <- c(10, 20, 30, 40, 50)
x
is.vector(x) #---> TRUE
```
```{r}
y <- c("Tom", "Dick", "Janey")
y
is.vector(y) #---> TRUE
```

```{r}
z <- c(1,2,3, "Tom", "Dick", "Janey")
is.vector(z) #---> TRUE
typeof(z) #----> character (Stored in memory as __ ATOMIC DATATYPE)
```

### The elements of a vector are called COMPONENTS.
### CLASS refers to the "Abstract type of the OBJECT in R".
### TYPE refers to "How the object is STORED in MEMORY by R.
### To get the class and type of x, use the following FUNCTIONS.

```{r}
class(x)
typeof(x) #------> Double- double precision floating point format.
```

### Character Vector

```{r}
y <- c("a", "b", "c")
y

class(y)
typeof(y)
```


### Logical Vector
```{r}
z <- c(TRUE, FALSE, FALSE, TRUE)
N <- c(FALSE, FALSE, TRUE, FALSE)
z
class(z)
typeof(z)
str(z)
```

### Vector of components of different types

```{r}
n <- c(1, 2, "a", "b", TRUE, FALSE)
n
class(n)
typeof(n)
str(n)
```

### Vector length (number of COMPONENTS)
```{r}
x <- c(10, 20, 30, 40, 50)
x
length(x)

y <- c("a", "b", "c")
y
length(y)
### we can add values to an existing vector
x <- c(x, 100:105)
y <- c(y,"d","e", "f", "g")
x
y
```

### If you nest two or more Vectors together,
### the result is ONE Vector ONLY.

```{r}
w <- c(c(1, 2, 3), 25)
w
str(w)
is.vector(w == 1)

k <- c(c(10,20,30), 25)
k
str(k)
is.vector(k == 1)

s <- c(c(2, 4, 6), c(10, 12, 14))
str(s)
is.vector(s == 1)
```

###################################################

########################################
# CREATING VECTORS WITH THE COLON OPERATOR (:)
########################################

### The COLON operator CREATES a sequence of INTEGERS
```{r}
x <- 1:5
x
typeof(x) #----> integer
```

### Get the CLASS(abstract type of object) and TYPE(stored in memory by R as_)
```{r}
class(x)#-----> class of object is integer
typeof(x)#----> type of vector is integer
```

### Create another vector with the same values, using the c() function
### Are Vectors with the SAME values, always Exactly IDENTICAL?
```{r}
x <- 1:5
y <- c(1, 2, 3, 4, 5)
x
y

x == y #---> TRUE TRUE TRUE TRUE TRUE
str(x) == str(y) #----> int[1:5] 1 2 3 4 5
                #----> num[1:5] 1 2 3 4 5
               #----> logical(0) 
              #----> NO
```
```{r}
class(y)#-----> class of object is NUMERIC
typeof(y)#----> type of VectorS are DOUBLE
```

```{r}
### Create SEQUENCES in Descending order
z <- 10:1
z
zyxwvut <- 26:20
zyxwvut 
```

```{r}
# Use the COLON Operator with VARIABLES
i <- 5
j <- 25

# Build a SEQUENCE from i to j
k <- (i:j)
k #-------> 5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 20 21 22 23 24 25
typeof(k)#-----> Atomic Data type is INTEGER

# BUILD A SEQUENCE from i to j By STEPS of 5
L <- c(i:j,steps=5)
L #----> 5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 20 21 22 23 24 25
### You can COMBINE the COLON Operator with the c() function
w <- c(3:9, 70, 49)
w
typeof(w)#----> Atomic data type is DOUBLE 
#-------------------------------------------------------------#
z <- c(1:10, 20:-5)
z
typeof(z)#----> The type is INTEGER.
        #----> Because the Vector contains ONLY int numbers.
Z_<- c(1:10, 20:-5,70)
typeof(Z_)#-----> Atomic data type is DOUBLE
```
###################################################

###################################
# CREATING VECTORS WITH THE rep() FUNCTION
###################################

### With rep() we CREATE Vectors with REPLICATED Values.

### You must specify the value and the number of times to replicate it.

# Replicate a vector of length one x(times).
```{r}
x <- rep(5, 3)
x
```
### we can do the same for strings and logical values

```{r}
y <- rep("Moms are cool", 6)
y
```
### You can combine rep() with c(),
### to replicate an entire vector (sequence of components)
### How many times will you replicate your Vector?

```{r}
z <- rep(c(1, 2, 3), 5)
z
```
### To replicate the vector element-by-element,
### we use the each argument.
```{r}
w <- rep(c(1, 2, 3), each = 5)
w
```

### You can combine rep() with the COLON Operator
```{r}
q <- rep(2:5, 3)
q
```
### To repeat the sequence element-by-element,
### use the each argument.
```{r}
s <- rep(2:5, each = 3)
s
```

### Other functions that REPLICATE values.
### The function numeric().
### It will RETURN a Vector of ZEROES.
```{r}
t <- numeric(14)
t
?numeric()
```
### The function character(), 
### will return a Vector of EMPTY Strings.

```{r}
k <- character(4)
k    #----> What application does a vector of empty strings have?
```

### The function logical() 
### will RETURN a Vector of FALSE Values.

```{r}
u <- logical(4)
u   #----> Application ?
```
########################################################


######################################
# CREATING VECTORS WITH THE seq() FUNCTION
######################################

### The seq() function creates sequences of REAL Numbers

### For this function you have to specify:
### The START value,the END value and the STEP (distance between components)
### the default step value is 1

```{r}
x <- seq(from=1, to=10, by=0.5)

x
```

### If we enter the arguments in this exact order (start, stop, step)
### we don't have to write the names

```{r}
y <- seq(1, 10, 0.5)

y
```

### other example

```{r}
z <- seq(0, 20, 2)
z
```

### one more example

```{r}
w <- seq(1, 10, 2)
w
```

### The sequence stops at 9, because the difference between 10 and 1
### is NOT a MULTIPLE of 2

### You can Create sequences in REVERSED order
### using a NEGATIVE step

# Countdown
```{r}
t <- seq(11, 1, -2)
t
```

### Instead of specifying the step, 
### you can specify the number of components
### in the sequence (the program will compute the step).
### To indicate the number of components USE the length argument

### create a sequence of 10 numbers between 1 and 5

```{r}
q <- seq(1, 5, length=10)

q
```
# the step is computed by dividing the difference 
# between the stop and start values
# to the number of components minus one

# in this case, the step is (5-1)/9 = 0.4444

### create a sequence of 100 numbers from -3 to 25

```{r}
#seq(_,_, length=_)
x <- seq(-3, 25, length = 100) #---> the step is (25+3)/99 = 0.282828

x
```

### You can combine the seq() function with c() and rep()
```{r}
y <- c(seq(1, 10, length=5), 30, 40)

y
```

### create a seq of vectors from 1 to 3, by 0.5, replicated three times.
```{r}
z <- rep(seq(1, 3, 0.5), 3)
z
```
#create a seq of vectors from 1to3 by 0.5, repeat each vector three times
```{r}
w <- rep(seq(1, 3, 0.5), each = 3)
w
```
###############################################################

##################################
# CREATING VECTORS OF RANDOM NUMBERS 
##################################

### Vectors of RANDOM NUMBERS are USEFUL in SIMULATIONS
### and other statistical applications

### Create a vector of discrete random numbers
### with the sample() function

### you must indicate the POPULATION
### and the number of values to be EXTRACTED (sample)

### Extract 12 random numbers from 1 to 100
### nCx (OUT OF 100, Choose 12 random numbers)

```{r}
n <- sample(100, 12)
n
```

# THE POPULATION CAN BE ANY VECTOR
```{r}
N<- sample(100:300,10)
N
```

### the population can be any vector
```{r}
h <- 100:300

H <- sample(h, 10)

H
```
### In these examples, the selection was made WITHOUT Replacement.
### So we got vectors of UNIQUE values.
### The replace option in the sample function() is set to FALSE by default.

### If we set it to TRUE, the selection will be made WITH Replacement

```{r}
p <- sample(30, 15, replace = TRUE)
p
```

### If the sample is greater than the population
### the Replace option MUST be SET to TRUE

### Generate/Extract 50 random numbers from 1 to 10

```{r}
m <- sample(10, 50, replace = TRUE)  
m
```

### If we don't allow the replacement, we get an error message
```{r}
M <- sample(10, 50, replace = FALSE)
M
```
# Generating Vectors of continuous random numbers
## using a few common statistical distributions

### To generate a vector of random numbers with a normal distribution
### we use the rnorm() function

### You must specify the number of values (TRIALS), mean and standard deviation.
### By default, mean is equal to 0 and standard deviation to 1.

```{r}
x <- rnorm(100, 10, 5)

x
```

# Generate 100 random numbers 
# with a Standardized Normal Distribution
# mean = 0, std = 1 
```{r}
x <- rnorm(100)
x
```
### to generate a vector of uniform random numbers
### we use the runif() function

# we must specify the number of trials and the range (lower and upper limit)
# The default range is 0-1
# lower limit = 0 and Upper limit = 1

# create a vector that contains 100 randon numbers 
# with a uniform distribution bewtween 1 and 5

```{r}
y <- runif(100, 1, 5)

y
```
### To generate a vector of EXPONENTIAL random numbers,
### Use the rexp() function.

### You must specify the number of trials,
### and the rate (1/mean).

# Generate a vector of exponential random numbers
# using the rexp () function
# the rate is 1/50
```{r}
z <- rexp(100, 1/50)

z
```
##################################################3

#####################
# CREATING EMPTY VECTORS 
#####################

### In some cases you need to INITIALIZE a vector before using it,
### by creating an empty vector.

### There are several ways to do this.

### Using an empty c() function

```{r}
x <- c()
x
typeof(x) #----> R stores it in memory as NULL
length(x) #----> 0
```

# Another way to create an empty vector 
# is to assign a NULL value to a variable
# using the NULL value

```{r}
y <- NULL
y
typeof(y) #----> NULL
is.vector(y) #----> FALSE
```
# A Third way to create an EMPTY VECTOR
# is using an empty vector() function

```{r}
z <- vector() #----> will return a logical empty vector
z
typeof(z) #---> stored in R memory as a "logical" vector
is.vector(z) #---> TRUE
is.logical(z) #---> TRUE
```

# TO ADD VALUES TO THE EMPTY VECTORS
# Simply concatenate the new values inside the empty vector
```{r}
x #--> NULL
x <- c(x, 1, 2)

x
typeof(x) #---> "double"

y #----> NULL
y <- c(y, 1, 2)

y
typeof(y) #---> "double"

z #----> logical(0)
z <- c(z, 1, 2)

z
typeof(z) #---> "double"
```
#####################################################

####################################
# INDEXING VECTORS USING NUMERIC INDICES 
####################################

### indexing a vector means accessing the vector components

### and the indices (subscripts) must be put between square brackets

```{r}
x <- c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100)
length(x)
### ACCESS one component (INDEXING)
x[2] #----> the 2nd component is (20)
```

### INDEX several components
# use COMBINE function c()
# we put them in a container
```{r}
x[c(1,3,8)] #----> 10 30 80
```
### INDEX components in order
```{r}
x[2:5] #----> 20 30 40 50

x[5:2] #----> 50 40 30 20

### combine the methods above

x[c(1:4, 7)] #---> 10 20 30 40 70
```

### ACCESS the last component (when we don't know the length)

```{r}
x[length(x)]
```
### we can put the indices in a separate vector
```{r}
n <- 3:5
x[n] #---> 30 40 50

```
### The NEGATIVE indices help us REMOVE a component

### remove one component
```{r}
x[-2]
```

### remove several components
### put the compnents you want to remove in a container
```{r}
x[c(-1, -3, -6)]

### remove components in order

x[-2:-4]

x[-4:-2]  # the order is not important here

### combine the methods above

# remove the first 5 components, and the 9th component
x[c(-1:-5, -9)]

# REMOVE THE LAST COMPONENT OF a VECTOR
# (when we don't know the length)
# Eliminate the last value in a vector 
x[-length(x)]

```

### You CANNOT mix negative and positive indices

```{r}
x[c(3, -7)] #-->ERROR: only zeros can be combined w/ negative indices

```
# To eliminate the 7th component in a Vector
# you can use this SYNTAX
```{r}
x[c(0, -7)]

```
### OR you can use this SYNTAX
### which has the same effect

```{r}
x[-7]
```

### You can add components to an existing vector
### (or edit components) using indices

```{r}
x <- c(7, 8, 9, 10)

x[5] <- 11

x

x[2] <- 100
x
```
### You cannot do this operation
# if the vector was not pre-defined
```{r}
p[1] <- 3 #----> Error in p[1] <- 3 : object 'p' not found
```
### You must define the vector first
# Create an EMPTY VECTOR 
# That we can add numbers to.
```{r}
p <- c() #---> NOW THE VECTOR P, DOES EXIST.
```
### then you can add components as you want

```{r}
p[1] <- 3

p[2] <- 7

p
```
### TO INSERT A COMPONENT into an existing vector
### you must "BREAK" the vector

```{r}
x <- c(9, 100, 11, 12)
x
```

### we want to insert 50 between 11 and 12

```{r}
# x <- c(9, 100, 11, 12)

x <- c(x[1:3], 50, x[4])

x
```
#####################################################3

####################################
# INDEXING VECTORS USING LOGICAL INDICES 
####################################

```{r}
x <- c(1, 2, 3, 4)
```
# We need to access the second and the third components
# Using logical indices

# Create a logical vector of the same length
```{r}
i <- c(FALSE, TRUE, TRUE, FALSE)

### Index the x vector

x[i] #---> 2 3 (TRUE TRUE)

### we can do this directly as well

x[c(FALSE, TRUE, TRUE, FALSE)]
```
###################################################

#######################
# NAMING VECTORS COMPONENTS 
#######################

### In some cases we might need to ASSIGN a NAME 
### To a/EACH component of a vector

### The vector below contains the SALARIES of THREE Employees
### Tom, Dick and Janey (in this order)

```{r}
salary <- c(4000, 2700, 3200)
```

### to add the employees names
# WE USE the names() function

```{r}
names(salary) <- c("Tom", "Dick", "Janey")
salary
```

### If a vector has names, 
### we can INDEX the components using the names

```{r}
salary["Tom"]
salary["Dick"]
salary["Janey"]
```
### to remove names we assign the NULL value to the names() function

```{r}
names(salary) <- NULL
salary
```

###########################################

#################
# FILTERING VECTORS 
#################

### VECTOR Filtering (or slicing) 
### is the operation of 
### selecting the components
### that meet one or more conditions

### the filtering conditions must be put between square brackets

```{r}
x <- c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100)
```

### FILTER the components that are GREATER than 45
```{r}
y <- x[x>45]
y
```
### FILTER the components that are smaller than 72

```{r}
y <- x[x<72]
y
```
### FILTER the components that are equal to 30

```{r}
y <- x[x==30]
y
```
### FILTER the components that are different from 30

```{r}
y <- x[x!=30]
y #---> 10 20 40 50 60 70 80 90 100
```
### select the components that are greater than 27
### AND lower than 83

```{r}
y <- x[x>27&x<83]
y #---> 30 40 50 60 70 80
```

### select the components that are lower than 35
### OR greater than 71

# THE VALUE MUST MEET EITHER CONDITON
```{r}
y <- x[x<35|x>71]
y #----> 10  20  30  80  90 100
```
### select the components that are lower than 60
### OR greater than 20

```{r}
y <- x[x<60|x>20]
y #---> ALL THE NUMBERS IN THE ARE CHOSEN
```

### select the components that are greater than 20
### AND different from 80

```{r}
y <- x[x>20&x!=80]
y
```
### select the components that are greater than 20
### OR different from 80

```{r}
y <- x[x>20|x!=80]
y
```
### To find out the INDICES of the components 
### that meet a condition
### USE THE  which() FUNCTION

### which components are EQUAL TO 40?

```{r}
which(x==40) #-----------> Index number 4

```
### which components are LESS THAN 54?
```{r}
which(x<54) #-----? Index number 1 2 3 4 5
```

# which components are GREATER THAN 43?
```{r}
which(x>43) #-----------------> 5 6 7 8 9 10
```

### of course, we can store these indices in a vector to use them later
```{r}
i <- which(x>43)
i
```

### to check whether a value can be found in a vector
### we can use the binary operator %in%

### it returns TRUE if the value is present

# CAN 40 BE FOUND IN THIS VECTOR?
```{r}
40 %in% x #-----> TRUE

# CAN 86 BE FOUND IN THIS VECTOR?
86 %in% x #----> FALSE
```
#######################################

##############
# all() AND any() FUNCTIONS
##############

### all() is a special function that checks whether all
### the components in a vector meet a certain condition

### the result of this FUNCTION is either TRUE or FALSE

```{r}
x <- c(1, 3, 5, 7, 9, 11, 13, 15)
```
### check whether all the components are greater than zero

```{r}
all(x>0) #----> TRUE
```
### check whether all the components are lower than 50
```{r}
all(x<50) #----> TRUE
```

### check whether all the components are lower than 8
```{r}
all(x<8)
```

### check whether all the components are greater than 10

```{r}
all(x>10)
```
### check whether all the components are greater than 0
### and lower than 10

```{r}
all(x>0&x<10) #-----> FALSE
```
### check whether all the components are greater than 0
### and lower than 35

```{r}
all(x>0&x<35) #----> TRUE
```

### check whether all the components are greater than 0
### or lower than 7

```{r}
all(x>0|x<7) #----> TRUE
```

### check whether all the components are greater than 4
### or lower than 7

```{r}
all(x>4|x<7) #----> TRUE
```

### check whether all the components are greater than 4
### or lower than 2

```{r}
all(x>4|x<2) #----> FALSE
```

### any() is a function that checks whether any component
### in a vector meets a certain condition

```{r}
x
```

### is there any component (any) greater than 4?
```{r}
any(x>4) #----> TRUE
```

### is there any component lower than than 0?
```{r}
any(x<0) #----> FALSE
```

### is there any component equal to 11?
```{r}
any(x==11) #----> TRUE
```

### is there any component equal to 4?
```{r}
any(x==4) #----> TRUE
```

# LETS TRY WITH TWO LOGICAL CONDITIONS
### is there any component lower than 10
### and greater than 3?

```{r}
any(x<10&x>3) #----> TRUE
```

### is there any component lower than 4
### and greater than 9?

```{r}
any(x<4&x>9) #----> TRUE
```
### is there any component lower than 4
### or greater than 9?

```{r}
any(x<4|x>9) #----> TRUE

### is there any component lower than 0
### or greater than 100?

any(x<0|x>100) # FALSE
```

############################################################

##################################
# HOW TO GET THE SUM AND PRODUCT OF VECTOR COMPONENTS
# BASIC STATISTICAL INDICATORS
# WE CAN ONLY DO THIS WITH NUMERIC VECTORS
##################################

```{r}
x <- c(2, 4, 6, 15, 17, 19)
```
### to add the components of a numeric vector

```{r}
sum(x)
```

### we can store the result in a new vector

```{r}
y <- sum(x)

y
```
### to multiply the components of a numeric vector
# prod function

```{r}
prod(x) #---> 232560
```
### to compute the cumulative sum

```{r}
cumsum(x) #----> 2  6 12 27 44 63
```

### to compute the cumulative product

```{r}
cumprod(x) #--> 2      8     48    720  12240 232560
```
### compute the mean of the components
# Computing STATISTICAL INDICATORS

```{r}
mean(x) #----> 10.5
```

### compute the median of the components

```{r}
median(x) #----> 10.5
```

### compute the variance of the components

```{r}
var(x) #---> 53.9
```

### compute the standard deviation of the components

```{r}
sd(x) #---> 7.341662
```


### get the summary statistics for a numeric vector
# this function provides us with the following information
```{r}
summary(x) #-----> Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
           #-----> 2.0     4.5    10.5    10.5    16.5    19.0 

fivenum(x) #----> 2.0  4.0 10.5 17.0 19.0 (Tukey's Summary)
```

##########################################################

#####################
# VECTORIZED OPERATIONS
#####################

### vectorized operations are operations that apply to all the components
### of a vector simultaneously (in parallel)

### This is very effective feature of R 
### that saves us from writing a lot of loops
### (vectorized operations are actually loops)
### They loop through the vector and 
### apply the operation to each component

### the result of a vectorized operation is a vector

### here's the main vectorized operations

### adding a vector to a scalar

```{r}
x <- c(1,2,3,4,5)

y <- x + 10 #---> This line of code takes the components of x, and adds 10 to all of them.
y #----> 11 12 13 14 15
```

### multiplying a vector by a scalar

```{r}
y <- x * 3 #---> multiples each component in our vector x by 3
y #----> 3  6  9 12 15
```

### computing the modulo values of the components
### (the remainders from dividing each component to a scalar)

```{r}
y <- x %% 2

y #---> 1 0 1 0 1
```

# Raising the vector components to a POWER
# Raise the component
```{r}
y <- x ^ 2

y #----> 1  4  9 16 25
```
### Exponentiate each component of a vector

```{r}
exp(x) #---> this computes 1^1 2^2 3^3 4^4 5^5
       #---> 1  4  9 16 25
```

### EXTRACT the SQUARE root
```{r}
sqrt(x) #---> 1.000000 1.414214 1.732051 2.000000 2.236068

x ^ (1/2) #---> 1.000000 1.414214 1.732051 2.000000 2.236068
```

### Extraction of any ROOT

```{r}
x ^ (1/3) #---> 1.000000 1.259921 1.442250 1.587401 1.709976
```

### Compute the NATURAL LOGARITHM

```{r}
log(x) #---> 0.0000000 0.6931472 1.0986123 1.3862944 1.6094379
```

### Compute the TRIGONOMETRICAL functions (sin, cos etc.)

```{r}
sin(x) #---> 0.8414710  0.9092974  0.1411200 -0.7568025 -0.9589243

cos(x) #--->  0.5403023 -0.4161468 -0.9899925 -0.6536436  0.2836622

tan(x) #---> 1.5574077 -2.1850399 -0.1425465  1.1578213 -3.3805150
```

### Compute the ABSOLUTE VALUE

```{r}
abs(x) #---> 1 2 3 4 5
```

### round the vector components

```{r}
z <- c(1.3, 3.7, 9.5)
```

# round function
```{r}
round(z)  #---> rounded components of the vector z
          #---> 1  4 10
```

### round to the closest LOWER interger
```{r}
floor(z) #---> 1 3 9
```

### round to the closest HIGHER interger
```{r}
ceiling(z) #----> 2  4 10
```

### round to a SPECIFIED Number of significant digits
### using the signif() function
### you have to specify the vector 
### and the number of digits

```{r}
w <- c(3.954, 7.200, 9.028)

signif(w, 1) #---> 4 7 9

signif(w, 2) #----> 4.0 7.2 9.0

signif(w, 3) #---> 3.95 7.20 9.03
```

### Compute the FACTORIAL of each element

```{r}
factorial(w) #----> 1   2   6  24 120
factorial(x) #----> 1   2   6  24 120
```
###############################################################

#######################
# MISSING VALUES IN VECTORS
#######################

### The missing (unknown) values are represented by the NA symbol

### Here's a vector that stores the salaries of five employees
```{r}
sal <- c(3500, 1800, 2400, NA, 2900)
```

### let's try to compute the average salary

```{r}
mean(sal)
```

### we must remove the missing value first
```{r}
sal2 <- na.omit(sal)
sal2 #----> 3500 1800 2400 2900
```

### Compute the mean

```{r}
mean(sal2) #---> 2650
```

### a simpler way to do the same
# Using the parameter na.rm, in the mean function

```{r}
mean(sal, na.rm = TRUE) #---> this option removes missing values automatically
```
### the missing values count for the vector length
```{r}
length(sal) #----> 5
```
### to identify the missing values we can use the function is.na()

# is a vector missing in the data, sal?
```{r}
y <- is.na(sal)

y #---->  FALSE FALSE FALSE  TRUE FALSE
```
#######################################################

###############
# SORTING A VECTOR
###############

```{r}
x <- c(6, 8, 3, 11, 5, 10, 2, 17)
```
#Apply sort function to get the numbers from lowest to highest 

```{r}
sort(x) #----> 2  3  5  6  8 10 11 17
```
### THE VECTOR ITSELF DOES NOT CHANGE 

### Check
```{r}
x #----> 6  8  3 11  5 10  2 17
```
### To get the order of indices we can use the order() function
### Only the order of INDICES changes 
### NOT THE VECTOR ITSELF

```{r}
#                2  3   5   6   8   10  11  17
order(x) #----> 7th 3rd 5th 1st 2nd 6th 4th 8th
```

### we can store the sorted values in an new vector
```{r}
y <- sort(x)
y
```
### to sort the components in decreasing order
# set the decreasing option to TRUE
```{r}
y <- sort(x, decreasing = TRUE)

y
```

### we can sort CHARACTER and LOGICAL vectors as well
```{r}
w <- c("z", "a", "n", "d")

# Components are sorted in alphabetical order
# By default the sort function sorts in ascending a-z order
sort(w)

# Set the sort function option to descending z-a order
sort(w, decreasing = TRUE)
```

### Logical Components
```{r}
q <- c(TRUE, FALSE, TRUE, FALSE)
```

# By default the sort function sorts logical values from FALSE to TRUE
```{r}
sort(q) #----> FALSE FALSE  TRUE  TRUE
```

# Set the option decreasing=TRUE to sort in descending order instead

```{r}
sort(q, decreasing = TRUE) #---> TRUE  TRUE FALSE FALSE
```
###################################################

#################
# MINIMA AND MAXIMA
#################

### to get the minimum and maximum values in a vector
### we use the min() and max() functions
```{r}
x <- c(4, 1, 7, 5, 12, 6)

min(x)

max(x)
```
### to determine the INDICES of the minimum and maximum values
### we use the which.min() and which.max() functions

#THE INDEX of the MINIMUM VALUE
```{r}
which.min(x) #----> 2
#THE INDEX of the MAXIMUM VALUE
which.max(x) #----> 5
```

### the functions pmin() and pmax() allow us to determine
### the minima and maxima from pairs of vector values
```{r}
y <- c(2, 9, 7, 6, 10, 14)
y #-----> 2  9  7  6  10  14
x #-----> 4  1  7  5  12  6

pmin(x, y) #----> 2  1  7  5 10  6 

pmax(x, y) #----> 4  9  7  6 12 14
```
##################################################

######################
# THE ifelse() FUNCTION
######################

### ifelse() is a simplified version of the if-else conditional statement
### it is applied to vectors

### the function syntax:

## ifelse(condition, action if true, action if false) ##

### condition - condition to be checked (tested)

### action if true - the action that the program must do if the condition is met

### action if false - the action that the program must do if the condition is NOT met

### the condition generates a logical vector
### that takes the value TRUE if the condition is satisfied
### and the value FALSE otherwise

### we will build an ifelse() function that checks whether the elements
### in a vector are even numbers
### if so, they will be DIVIDED by TWO, if ELSE they will be copied
```{r}
x <- c(6, 10, 9, 5, 20, 7, 16)

y <- ifelse(x%%2 == 0, x/2, x)

y #----> 3  5  9  5 10  7  8
```
### create an ifelse() function that multiplies the component by 5
### if it is lower than or equal to 12 and by 3 if it is greater than 12

```{r}
x <- c(2, 6, 15, 10, 20, 14)

y <-ifelse(x<=12, x * 5, x * 3) #-->ifelse(condition=TRUE, DO action1, otherwise DO action2)

y #---> 10 30 45 50 60 42
```

### Create another ifelse() function that does the following:
### if the component is greater than or equal to zero, it extracts the square root
### otherwise, it changes the component sign and then extracts the square root

```{r}
x <- c(25, -36, 100, 0, -1, 49, 4, -68)

y <- ifelse(x>=0, sqrt(x), sqrt(-x))

y #---> 5.000000  6.000000 10.000000  0.000000  1.000000  7.000000  2.000000  8.246211

# Warning messages---> Because the program computed the SQRT of negative numbers
# Before changing the SIGNS
"1: In sqrt(x) : NaNs produced
 2: In sqrt(-x) : NaNs produced"
### TO REMOVE WARNINGS, we can do as follows

y <- sqrt(ifelse(x>=0, x, -x)) #--> introduce the ifelse function within the sqrt function
### IF the COMPONENT is GREATER than or EQUAL to zero, the COMPONENT is COPIED
### IF it is LOWER than zero, the SIGN is CHANGED

y #---> SAME AS LINE #? above
```

### one last example: an ifelse() function that returns 1
### if the square of the component is greater than 100 and 0 otherwise
```{r}
x <- c(15, 6, 10, 30, 4)

y <- ifelse(x^2>100, 1, 0) #--> if the Square of the COMPONENT is greater than 100,
                           #--> the function returns 1, IF ELSE it returns 0
#Lets see
y #-----> 1 0 0 1 0

```
#######################################################





###########################
# ADDING AND MULTIPLYING VECTORS
###########################

### the vectors of the same length can be added and multiplied element-wise

```{r}
x <- c(2, 4, 6, 8)

y <- c(10, 20, 30, 40)
```

```{r}
z <- x + y

z
```

```{r}
x - y #---> -8 -16 -24 -32

x * y #---> 20  80 180 320

x / y #---> 0.2 0.2 0.2 0.2

y / x #---> 5 5 5 5
```

# Compute the MODULO 
# Remainder of dividing each component of x by each component of y
```{r}

x %% y #----> 2 4 6 8

y %% x #---> 0 0 0 0

```

### if the vectors have different lengths
### the shorter vector is RECYCLED

```{r}
x <- c(1, 2, 3, 4)

y <- c(9, 10)

x + y #---> 10 12 12 14

### y was recycled to (9, 10, 9, 10)

#compute multiplication
x * y #----> 9 20 27 40

#compute MODULO
y %% x #---> 0 0 0 2
```
##################################################

#######################
# TESTING VECTOR EQUALITY
#######################

### the comparisons between vectors are vectorized operations
### the comparison is done element-wise

```{r}
x <- c(7, 10, 3, 9, 6)

y <- c(2, 14, 5, 15, 6)
```
### the operation below returns TRUE for the pairs of components
### where x is greater than y and FALSE for the pairs of components
### wherey y is greater than x
```{r}
x > y #----> TRUE FALSE FALSE FALSE FALSE

```

### the operation below returns TRUE for the pairs of components
### where y is greater than x and FALSE for the pairs of components
### wherey x is greater than y

```{r}
y > x #----> FALSE  TRUE  TRUE  TRUE FALSE

### the operation below returns TRUE for the pairs of components
### where x is equal to y

y == x #----> FALSE FALSE FALSE FALSE  TRUE
```

### To check whether two vectors are equal
### (all the components are the same)
```{r}
x <- c(2, 4, 6)

y <- c(2, 4, 6)

all(y == x) #----> FALSE
```
### the function all() can be used for other comparisons too

```{r}
x <- c(7, 8, 9)

y <- c(2, 4, 6)
```

### check whether all components in x are greater than 
### their corresponding components in y

```{r}
all(x>y) #----> TRUE

# conversely

all(y>x) #-----> FALSE
```

### the function any() can also be used for comparisons

```{r}
x <- c(4, 5, 6, 8)

y <- c(2, 7, 10, 8)
```

### check whether at least one component in x is greater than 
### its corresponding component in y
```{r}
any(x>y) #----> TRUE

# conversely

any(y>x) #----> TRUE

# check whether at least one component in x is equal to 
# its corresponding component in y

any(x==y) #---> TRUE
```
### another function that allows us to test vector equality
### is identical()

```{r}
x <- c(1, 2, 3)

y <- c(1, 2, 3)

z <- c(1, 2, 5)

identical(x, y) #----> TRUE

identical(x, z) #----> FALSE
```
### another function is all.equal()
### if the vectors are identical, it returns TRUE
### if not, it returns the MEAN DIFFERENCE BETWEEN COMPONENTS
```{r}
all.equal(x, y) #----> TRUE

all.equal(x, z)  # mean difference is 0.6667: (5-3)/3
```

### EQUAL is NOT Necessarily IDENTICAL
```{r}
x <- c(4, 5, 6) 

y <- 4:6

identical(x, y) #----> FALSE
typeof(x) #--> "double"
typeof(y) #--> "integer
### these vectors are not identical, because they are of different types
typeof(x)
typeof(y)

### however, the all.equal() function returns TRUE

all.equal(x, y) #----> TRUE

### that's because all.equal() evaluates the "near identity"
### it is interested only in the components values
### and not in the vectors type
```
#######################################################
##################
# VECTOR CORRELATION
##################

### we can use the Pearson correlation coefficient
### to assess the correlation between two numeric vectors

### the vectors must have the same length

### the vector of the employees' annual salaries (in thousands)

```{r}
sal <- c(80, 66, 42, 40, 75, 82, 54, 44, 85, 83)
```
### the vector of the employees' education (in years of school completed)
```{r}
edu <- c(15, 18, 12, 12, 16, 17, 13, 12, 17, 16)
```
### to compute the correlation we use the cor.test() function

```{r}
cor.test(sal, edu) #----> 0.858553 
```
### Pearson's product-moment correlation

### data:  sal and edu
### t = 4.7362, df = 8, p-value = 0.001471
### alternative hypothesis: true correlation is not equal to 0
### 95 percent confidence interval:
### 0.4982815 0.9659942
### sample estimates:
###   cor
### 0.858553') 

#########################################################




